Dominance Regions for Rank Two Cluster Algebras

### 2022-05-29 dominance_regions.tex
Added some references at the start of section 3, the statement probably needs work though.
Going to sleep

  Fix integers $b,c>0$.
  The cluster algebra $\cA(b,c)$ is the $\ZZ$-subalgebra of $\QQ(x_0,x_1)$ generated by the \emph{cluster variables} $x_m$, $m\in\ZZ$, defined recursively by
  \begin{equation}
    x_{m-1}x_{m+1}=\begin{cases} x_m^b+1 & \text{if $m$ is even;}\\ x_m^c+1 & \text{if $m$ is odd.} \end{cases}
  \end{equation}
  By the Laurent Phenomenon \cite{FZ02}, each $x_m$ is actually an element of $\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for any $k\in\ZZ$.
  Moreover, these Laurent polynomials are known to have positive coefficients \cite{GHKK18,LLZ14,LS15}.

@@

  \begin{lemma}
    \label{le:limits}
    We have
    \[\lim_{i\to\infty} \frac{u_i^-}{u_{i-1}^+}=\frac{bc+\sqrt{bc(bc-4)}}{2b}=\lim_{i\to-\infty} \frac{u_{i-1}^-}{u_i^+};\]
    \[\lim_{i\to\infty} \frac{u_{i-1}^-}{u_i^+}=\frac{bc-\sqrt{bc(bc-4)}}{2b}=\lim_{i\to-\infty} \frac{u_i^-}{u_{i-1}^+}.\]
    Moreover, the limits in the first line converge monotonically from above and the limits in the second line converge monotonically from below.
  \end{lemma}
  \begin{remark}
    The analogous limits with $\varepsilon=+$ and $\varepsilon=-$ reversed are obtained from these by interchanging the roles of $b$ and $c$.
  \end{remark}
  \begin{proof}
    We first compute the limits for arbitrary positive real numbers $b$ and $c$ with $bc>4$; the remaining case will then follow by taking the limit as $bc$ approaches $4$.

@@

  We begin by studying transformations of $\RR^2$ which determine the change of $\bfg$-vectors when expanding a pointed expression of the form~\eqref{eq:pointed} in terms of a cluster $\{x_k,x_{k+1}\}$.
  Write $\phi_0:\RR^2\to\RR^2$ for the identity map and define piecewise-linear maps $\phi_{\pm 1}:\RR^2\to\RR^2$ as follows:
  \begin{equation}
    \label{eq:forward mutation 1}
    \phi_1(\lambda)
    :=
    \begin{cases} 
      (-\lambda_0,c\lambda_0+\lambda_1) & \text{if $\lambda_0 \ge 0$;}\\
      (-\lambda_0,\lambda_1) & \text{if $\lambda_0 < 0$;}
    \end{cases}
    \qquad
    \phi_{-1}(\lambda)
    :=
    \begin{cases} 
      (\lambda_0,-\lambda_1) & \text{if $\lambda_1 \ge 0$;}\\
      (\lambda_0+b\lambda_1,-\lambda_1) & \text{if $\lambda_1 < 0$.}
    \end{cases}
  \end{equation}
  For $k\in\ZZ$ with $|k|>1$, define piecewise-linear maps
  \[\phi_k
    :=
    \begin{cases}
      (\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j$;}\\
      \phi_1(\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j+1$.}\\
    \end{cases}
  \]
  These determine the dominance region as follows.
  \begin{definition}
    \label{def:dominance}
    For $\lambda\in\ZZ^2$ and $k\in\ZZ$, define cones 
    \[\cC_k(\lambda)
      :=
      \begin{cases}
        \{(\lambda_0-r,\lambda_1+s):r,s\in\RR_{\ge0}\} & \text{if $k=2j$;}\\
        \{(\lambda_0+r,\lambda_1-s):r,s\in\RR_{\ge0}\} & \text{if $k=2j+1$.}
      \end{cases}
    \]
    The \emph{dominance region} $\cP_\lambda$ is the intersection  $\bigcap_{k\in\ZZ}\phi_k^{-1}\cC_k(\phi_k\lambda)$.
    When $\mu\in\cP_\lambda$, we say \emph{$\lambda$ dominates $\mu$}.
  \end{definition}

@@

  By an eigenvector of a piecewise-linear map $\phi$, we will mean a vector $\lambda$ such that for each $n>0$, there exists a scalar $\nu_n$ so that $\phi^n(\lambda)=\nu_n\lambda$.
  \begin{lemma}
    Any nonzero eigenvector of $\phi_2$ is a positive multiple of one of the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$.
  \end{lemma}
  \begin{proof}
    First observe that the equation $\phi_2^n(\lambda)=\nu_n\lambda$ cannot be satisfied for all $n>0$ unless $\lambda_0\ge 0$ and $c\lambda_0+\lambda_1\ge 0$.
    In this region, $\phi_2$ is linear with eigenvalue $\nu$ satisfying $\nu^2-(bc-2)\nu+1=0$, i.e. $\nu=\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}$.
    We thus require 
    \[\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_0=(bc-1)\lambda_0+b\lambda_1 \qquad\text{and}\qquad \frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_1= -c\lambda_0-\lambda_1,\]
    or equivalently
    \[\frac{-bc\pm\sqrt{bc(bc-4)}}{2b}\lambda_0=\lambda_1 \qquad\text{and}\qquad \frac{-bc\mp\sqrt{bc(bc-4)}}{2c}\lambda_1=\lambda_0.\]
    As these represent the same relationship, the result immediately follows by inspection.
  \end{proof}

  %\begin{lemma}
  %  The vectors $\big(2b, -bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2b, -bc-\sqrt{bc(bc-4)}\big)$) and $\big(2c, bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2c, bc-\sqrt{bc(bc-4)}\big)$) are perpendicular.
  %\end{lemma}
  %\begin{proof}
  %  This is immediate from the equality $\big(-bc+\sqrt{bc(bc-4)}\big)\big(bc+\sqrt{bc(bc-4)}\big)=-4bc$.
  %\end{proof}
  \begin{lemma}
    \label{le:imaginary stability}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have $\phi_{2j}(\lambda)\in\cI$.
  \end{lemma}
  \begin{proof}
    Since $\phi_{2j}=\phi_2^j$, the result follows from the cases $j=1$.
    But this is immediate for $\phi_2$ since the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$ are eigenvectors and $\phi_2$ is linear in $\cI$.
  \end{proof}

  It will be useful to have explicit expressions for $\phi_k(\lambda)$ for $\lambda\in\cI$.
  \begin{lemma}
    \label{le:imaginary transformations}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have
    \begin{align*}
      \phi_{2j}(\lambda)&=(u_{2j+1}^+\lambda_0+u_{2j}^+\lambda_1,-u_{2j}^-\lambda_0-u_{2j-1}^-\lambda_1);\\
      \phi_{2j+1}(\lambda)&=(-u_{2j+1}^+\lambda_0-u_{2j}^+\lambda_1,u_{2j+2}^-\lambda_0+u_{2j+1}^-\lambda_1).
    \end{align*}
  \end{lemma}
  \begin{proof}
    We work by induction on $j$, the case $j=0$ being clear from the definitions.
    For $\lambda\in\cI$, the action of $\phi_2$ from \eqref{eq:forward two step mutation} can be rewritten as
    \[\phi_2(\lambda)=(u_3^-\lambda_0+u_2^+\lambda_1,-u_2^-\lambda_0-u_1^+\lambda_1).\]
    Therefore
    \begin{align*}
      \phi_{2j+2}(\lambda)&=\phi_2\phi_{2j}(\lambda)\\
      &=\big( (u_3^- u_{2j+1}^+-u_2^+u_{2j}^-)\lambda_0+(u_3^-u_{2j}^+-u_2^+u_{2j-1}^-)\lambda_1,-(u_2^-u_{2j+1}^+-u_1^+u_{2j}^-)\lambda_0-(u_2^-u_{2j}^+-u_1^+u_{2j-1}^-)\lambda_1 \big)\\
      &=(u_{2j+3}^+\lambda_0+u_{2j+2}^+\lambda_1, -u_{2j+2}^-\lambda_0-u_{2j+1}^-\lambda_1),
    \end{align*}
    as desired, where the last equality uses \eqref{eq:long Chebyshev recursion} with $i=2j+1$ and $\ell=2$.

>>>>

  Fix integers $b,c>0$.
  The cluster algebra $\cA(b,c)$ is the $\ZZ$-subalgebra of $\QQ(x_0,x_1)$ generated by the \emph{cluster variables} $x_m$, $m\in\ZZ$, defined recursively by
  \[
    x_{m-1}x_{m+1}=\begin{cases} x_m^b+1 & \text{if $m$ is even;}\\ x_m^c+1 & \text{if $m$ is odd.} \end{cases}
  \]
  By the Laurent Phenomenon \cite{FZ02}, each $x_m$ is actually an element of $\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for any $k\in\ZZ$.
  Moreover, these Laurent polynomials are known to have positive coefficients \cite{GHKK18,LLZ14,LS15}.

@@

  \begin{lemma}
    \label{le:limits}
    We have
    \[\lim_{i\to\infty} \frac{u_i^-}{u_{i-1}^+}=\frac{bc+\sqrt{bc(bc-4)}}{2b}=\lim_{i\to-\infty} \frac{u_{i-1}^-}{u_i^+};\]
    \[\lim_{i\to\infty} \frac{u_{i-1}^-}{u_i^+}=\frac{bc-\sqrt{bc(bc-4)}}{2b}=\lim_{i\to-\infty} \frac{u_i^-}{u_{i-1}^+}.\]
    Moreover, the limits in the first line converge monotonically from above and the limits in the second line converge monotonically from below.
  \end{lemma}
  \begin{remark}
    The analogous limits with $\varepsilon=+$ and $\varepsilon=-$ reversed are obtained from these by interchanging the roles of $b$ and $c$.
  \end{remark}
  \begin{proof}
    When $bc=4$ it is easy to compute closed formulas for $u_i^\varepsilon$ and the claimed limits follow; we thus concentrate on the case $bc>4$.

@@

  We begin by studying transformations of $\RR^2$ which determine the change of $\bfg$-vectors when expanding an expression of the form~\eqref{eq:pointed} in terms of a cluster $\{x_k,x_{k+1}\}$.
  This adapts the notation from \cite[Definition 2.1.4]{Qin19} to our setting, see also \cite[Definition 4.1]{Rea14}.

  Write $\phi_0:\RR^2\to\RR^2$ for the identity map and define piecewise-linear maps $\phi_{\pm 1}:\RR^2\to\RR^2$ as follows:
  \begin{equation}
    \label{eq:forward mutation 1}
    \phi_1(\lambda)
    :=
    \begin{cases} 
      (-\lambda_0,c\lambda_0+\lambda_1) & \text{if $\lambda_0 \ge 0$;}\\
      (-\lambda_0,\lambda_1) & \text{if $\lambda_0 < 0$;}
    \end{cases}
    \qquad
    \phi_{-1}(\lambda)
    :=
    \begin{cases} 
      (\lambda_0,-\lambda_1) & \text{if $\lambda_1 \ge 0$;}\\
      (\lambda_0+b\lambda_1,-\lambda_1) & \text{if $\lambda_1 < 0$.}
    \end{cases}
  \end{equation}
  For $k\in\ZZ$ with $|k|>1$, define piecewise-linear maps
  \[\phi_k
    :=
    \begin{cases}
      (\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j$;}\\
      \phi_1(\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j+1$.}\\
    \end{cases}
  \]
  These determine the dominance region as we paraphrase from \cite[Section 3.1]{Qin19}.
  \begin{definition}
    \label{def:dominance}
    For $\lambda=(\lambda_0,\lambda_1)\in\ZZ^2$ and $k\in\ZZ$, define cones 
    \[\cC_k(\lambda)
      :=
      \begin{cases}
        \{(\lambda_0-r,\lambda_1+s):r,s\in\RR_{\ge0}\} & \text{if $k=2j$;}\\
        \{(\lambda_0+r,\lambda_1-s):r,s\in\RR_{\ge0}\} & \text{if $k=2j+1$.}
      \end{cases}
    \]
    The \emph{dominance region} $\cP_\lambda$ is the intersection  $\bigcap_{k\in\ZZ}\phi_k^{-1}\cC_k(\phi_k\lambda)$.
    When $\mu\in\cP_\lambda$, we say \emph{$\lambda$ dominates $\mu$}.
  \end{definition}

@@

  By an eigenvector of a piecewise-linear map $\phi$, we will mean a vector $\lambda$ such that for each $n>0$, there exists a positive scalar $\nu$ so that $\phi(\lambda)=\nu\lambda$.
  \begin{lemma}
    Any nonzero eigenvector of $\phi_2$ is a positive multiple of one of the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$.
  \end{lemma}
  \begin{proof}
    First observe that, by equation \eqref{eq:forward two step mutation}, the equation $\phi_2(\lambda)=\nu\lambda$ cannot be satisfied with a positive $\nu$ unless $\lambda_0\ge 0$ and $c\lambda_0+\lambda_1\ge 0$.
    In this region, $\phi_2$ is linear with eigenvalue $\nu$ satisfying $\nu^2-(bc-2)\nu+1=0$, i.e.~$\nu=\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}$.
    We thus require 
    \[\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_0=(bc-1)\lambda_0+b\lambda_1 \qquad\text{and}\qquad \frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_1= -c\lambda_0-\lambda_1,\]
    or equivalently
    \[\frac{-bc\pm\sqrt{bc(bc-4)}}{2b}\lambda_0=\lambda_1 \qquad\text{and}\qquad \frac{-bc\mp\sqrt{bc(bc-4)}}{2c}\lambda_1=\lambda_0.\]
    As these represent the same relationship, the result immediately follows by inspection.
  \end{proof}

  %\begin{lemma}
  %  The vectors $\big(2b, -bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2b, -bc-\sqrt{bc(bc-4)}\big)$) and $\big(2c, bc+\sqrt{bc(bc-4)}\big)$ (resp. $\big(2c, bc-\sqrt{bc(bc-4)}\big)$) are perpendicular.
  %\end{lemma}
  %\begin{proof}
  %  This is immediate from the equality $\big(-bc+\sqrt{bc(bc-4)}\big)\big(bc+\sqrt{bc(bc-4)}\big)=-4bc$.
  %\end{proof}
  Observe that the imaginary cone $\cI$ is spanned by the eigenvectors of $\phi_{2}$ and that $\phi_{2}$ is linear in $\cI$.
  \begin{lemma}
    \label{le:imaginary stability}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have $\phi_{2j}(\lambda)\in\cI$.
  \end{lemma}
  \begin{proof}
    Since $\phi_{2j}=\phi_2^j$, the result follows from the case $j=1$ which is immediate.
  \end{proof}

  It will be useful to have explicit expressions for $\phi_k(\lambda)$ for $\lambda\in\cI$.
  \begin{lemma}
    \label{le:imaginary transformations}
    For $j\in\ZZ$ and $\lambda\in\cI$, we have
    \begin{align*}
      \phi_{2j}(\lambda)&=(u_{2j+1}^-\lambda_0+u_{2j}^+\lambda_1,-u_{2j}^-\lambda_0-u_{2j-1}^+\lambda_1);\\
      \phi_{2j+1}(\lambda)&=(-u_{2j+1}^-\lambda_0-u_{2j}^+\lambda_1,u_{2j+2}^-\lambda_0+u_{2j+1}^+\lambda_1).
    \end{align*}
  \end{lemma}
  \begin{proof}
    We work by induction on $j$, the case $j=0$ being clear from the definitions.
    For $\lambda\in\cI$, the action of $\phi_2$ from \eqref{eq:forward two step mutation} can be rewritten as
    \[\phi_2(\lambda)=(u_3^-\lambda_0+u_2^+\lambda_1,-u_2^-\lambda_0-u_1^+\lambda_1).\]
    Therefore
    \begin{align*}
      \phi_{2j+2}(\lambda)&=\phi_2\phi_{2j}(\lambda)\\
      &=\big( (u_3^- u_{2j+1}^+-u_2^+u_{2j}^-)\lambda_0+(u_3^-u_{2j}^+-u_2^+u_{2j-1}^-)\lambda_1,-(u_2^-u_{2j+1}^+-u_1^+u_{2j}^-)\lambda_0-(u_2^-u_{2j}^+-u_1^+u_{2j-1}^-)\lambda_1 \big)\\
      &=(u_{2j+3}^+\lambda_0+u_{2j+2}^+\lambda_1, -u_{2j+2}^-\lambda_0-u_{2j+1}^-\lambda_1),
    \end{align*}
    as desired, where the last equality uses \eqref{eq:long Chebyshev recursion} with $i=2j+1$ and $\ell=2$.
